Databases are an effective and efficient method for storage and access of mass-spectrometry data

### 2025-03-11 manuscript/manuscript.Rmd
Nearly completed first draft.

mz5's documentation was remarkably sparse, especially considering it was one of the earliest mzML formats and is supported by Proteowizard. Crucially, the original paper [@Wilhelm2012] contains links to a website (https://software.steenlab.org/mz5) which currently returns an HTTP error 500. A Python library (pymz5) exists but requires an old version of Python (2.7 or 3.2) and hasn't been updated in 12 years and is predominately a simple fork of h5py [@ref] with three mz5-specific commits on top. Most problematically, we were unable to determine how mz5 stores precursor *m/z* ratios, making the fragment and precursor searches impossible. This was largely due to the variable-length nested compound structures mz5 that are not supported in all APIs, e.g. Java (@Bhamber2021).

@@

We then used custom code to convert the mzML files into SQLite and DuckDB databases using a simple schema for MS1 and MS2 data. The MS1 table consisted simply of fields for filename, scan index, retention time, *m/z* ratio, and intensity. The MS2 table consisted of the same fields except that the *m/z* column was separated into precursor and fragment *m/z*. Although we did not extend these databases to include the metadata associated with each file and scan, the logical framework could be easily extended in future work and the metadata typically represents a small fraction of the total space within the file, allowing us to make reasonable comparisons about file size between the databases and the metadata-rich other file types. We also converted each file's MS1 and MS2 table into Parquet representations for comparison using the same field/column schema.

@@

```{r singlefile_fig}
all_timings <- read_csv("data/singlefile_times.csv") %>% 
  mutate(method=factor(method, levels=method_names)) %>%
  mutate(query=factor(query, levels=c("ms1_spec", "chrom", "rtrange",
                                      "ms2_spec", "premz", "fragmz"),
                      labels=c("MS1 scan", "Chromatogram", "RT Range",
                               "MS2 scan", "Precursor search",
                               "Fragment search")))
all_sizes <- read_csv("data/file_sizes.csv") %>%
  mutate(method=factor(method, levels=method_names))

@@

*Figure XX: Query time for the six data extraction methods and the associated file sizes for all 13 methods explored in this paper. The left six panels show boxplots representing the time required in seconds to extract an MS1 scan, an ion chromatogram, all data within a retention time range, an MS2 scan, the fragments of a specified precursor, and all precursors with a specified fragment. The error in the boxplot is composed of timing information for 10 repeated queries, each of a different target scan number, retention time (RT), or m/z. The right panel shows a barplot of the size on disk in megabytes (MB) occupied by each file type.*

@@

@@

As the gap between data scientist and mass spectrometrist continues to narrow, mass spectrometry (MS) data formats should facilitate this convergence.

Large gaps etc. accomplished by extremely poor documentation that makes it difficult to do anything beyond the original designer's intent, typically with data structures that are not familiar to the MS expert.

Not a single parser was without flaws - all were either slow, poorly documented, or had highly limited functionality

Several fundamental inefficiencies identified:
1. Looping over each scan within a file
2. Looping over each file within a dataset
3. Decoding mz/int information each time if filtering on these fields
4. RTs are not accessible outside of looping over each scan for RT subsetting
5. Scan ids are not accessible outside of looping over each scan for scan extraction

Querying things by scan number is dumb.
1) Different indices for pyopenms vs pymzml vs pyteomics
2) idx = scan_num only works for consecutive scans (no multi-experiment, no polarity switching, no filtering)
3) Only USEFUL when you're parsing the entire file! Nobody knows what scan number 976 corresponds to, making extraction by number useless

DuckDB's zonemaps operate similarly to mzDB's bounding boxes
DuckDB also has automatic compression detection (e.g. the delta encoding of mz5)

Unlikely that databases replace the original files so this is meant to be an auxiliary data structure **alongside** the .raw or .mzML files, not a substitute.

Databases place a larger burden on the initial builder/designer and much less on the end user

Better compression algorithms and fancier data structures can easily improve upon the text-based XML in both size on disk and speed of access but also tend to accumulate significant technical debt. Alternatives must therefore compete not only in computational efficacy but also in simplicity to allow widespread adoption, a feat thus far unachieved.

Ultimately, the design decision for mass spectrometry data format will likely result from a variety of factors, most crucially 1) initial vendor type, 2) programming language of the developer, 3) types of MS data included (MS1-only vs MS2+, metadata), 4) whether the entire file will be processed or only a subset (subsetting doesn't matter if the whole file is going to be investigated anyway),

@@

>>>>

mz5's documentation was remarkably sparse, especially considering it was one of the earliest mzML formats and is supported by Proteowizard. Crucially, the original paper [@Wilhelm2012] contains links to a website (https://software.steenlab.org/mz5) which currently returns an HTTP error 500. A Python library (pymz5) exists but requires an old version of Python (2.7 or 3.2) and hasn't been updated in 12 years and is predominately a simple fork of h5py [@Collette2017] with three mz5-specific commits on top. Most problematically, we were unable to determine how mz5 stores precursor *m/z* ratios, making the fragment and precursor searches impossible. This was largely due to the variable-length nested compound structures mz5 that are not supported in all APIs, e.g. Java (@Bhamber2021).

@@

We then used custom code to convert the mzML files into SQLite and DuckDB databases using a simple schema for full scan (MS1) and MS/MS (MS2) data. The MS1 table consisted simply of fields for filename, scan index, retention time, *m/z* ratio, and intensity. The MS2 table consisted of the same fields except that the *m/z* column was separated into precursor and fragment *m/z*. Although we did not extend these databases to include the metadata associated with each file and scan, the logical framework could be easily extended in future work and the metadata typically represents a small fraction of the total space within the file, allowing us to make reasonable comparisons about file size between the databases and the metadata-rich other file types. We also converted each file's MS1 and MS2 table into Parquet representations for comparison using the same field/column schema.

@@

```{r singlefile_fig}
all_timings <- read_csv("data/singlefile_times.csv") %>% 
  mutate(method=factor(method, levels=method_names)) %>%
  mutate(query=factor(query, levels=c("ms1_spec", "chrom", "rtrange",
                                      "ms2_spec", "premz", "fragmz"),
                      labels=c("Full scan", "Chromatogram", "RT Range",
                               "MS/MS scan", "Precursor search",
                               "Fragment search")))
all_sizes <- read_csv("data/file_sizes.csv") %>%
  mutate(method=factor(method, levels=method_names))

@@

*Figure XX: Query time for the six data extraction methods and the associated file sizes for all 13 methods explored in this paper. The left six panels show boxplots representing the time required in seconds to extract an full scan spectrum, an ion chromatogram, all data within a retention time range, an MS/MS scan, the fragments of a specified precursor, and all precursors with a specified fragment. The error in the boxplot is composed of timing information for 10 repeated queries, each of a different target scan number, retention time (RT), or m/z. The right panel shows a barplot of the size on disk in megabytes (MB) occupied by each file type.*

@@

*Figure XX: blah*

@@

As the gap between data scientist and mass spectrometrist continues to narrow, mass spectrometry (MS) data formats should facilitate this convergence. Instead, MS software remains relatively opaque. Documentation is sparse and data structures are complex, resulting in a landscape that is essentially restricted to the original developer's intent. A particular pain point is the way in which MS data is stored because current methods must make trade-offs between simplicity, size, and speed. When we explored the wide range of MS file readers, we found every method had flaws that interfered with widespread adoption. The mzML file type appears to represent low-hanging fruit, with its XML-based structure that sacrifices speed and size in favor of clarity, but no alternative has yet reached a large audience of active users. Fifteen years of active competition continue to favor the highly explicit format, likely because users are leery of incomprehensible alternatives with no guarantee of continued maintenance.

Certain methods are clear winners for individual use cases but all of the existing formats failed to perform well at the full suite of exploratory data analysis tasks we attempted. Scan extraction is perhaps the most widely used query, especially for proteomics, and as a result has been extensively optimized. Here, MZA was blazingly fast thanks to the decision to index in the HDF5 file by scan number. The mzML files used here did have a precompiled index that should have made scan extraction highly efficient but this appeared to be mitigated for pyopenms and pyteomics due to their long initialization times, though pymzml performed very well at this task and the other two mzML methods were much faster after initialization (Supp Fig XX).

However, performant scan-based methods struggled significantly with chromatogram extraction because these axes are inherently orthogonal to each other. Here, mzDB performed quite well and was competitive with the specifically optimized mzTree and mzMD formats and pyopenms performed incredibly well when extracting multiple chromatograms via its 2DPeak method. Chromatograms (and retention time range queries) are of course only relevant for chromatography-based workflows but this type of analysis has become increasingly popular and we expect it to continue to do so, making ion extraction increasingly important. None of the existing MS formats we tested performed very well on MS/MS data, despite the growing availability of fragmentation data, though our use-case is oriented more towards exploration instead of comprehensive analysis.

Finally, a complexity penalty must be noted for formats and packages requiring complex installation procedures. pyopenms appears to be the worst offender here, with its bindings to the OpenMS C++ libraries requiring us to step back to Python 3.11 and numpy 1.26 to successfully access the data on our setup and required direct input from the maintainers. The Java applications for MzTree and, more egregiously mzMD, provided essentially zero documentation on installation consisting of a single README file without an intact HTTP request example and have not seen updates in years. Similarly, mzDB files were difficult to parse due to its opaque SQLite schema and use of the BLOB encoding type that again lacked examples or documentation outside of Java and had to be deduced iteratively. Of course, these methods eventually provided enough information that they could be parsed unlike Aird, toffee, and UIMF.

### Timing comparison to existing literature

Novel formats are typically proposed with timing and sizing information, but the inconsistency of what's being queried makes it difficult to directly compare across the literature. However, we mostly obtained results in line with those reported elsewhere where other intercomparisons have been performed and report here the widest set of comparisons between MS formats to our knowledge.

@Wilhelm2012 performed comparisons between mz5 and mzML with results indicating that mz5 was three times faster than mzML parsers. Their values of 0.16 seconds per million *m/z*/intensity arrays correspond to an estimated query time of 0.13 milliseconds which is faster than our measured 3 milliseconds, though our mz5 file was approximately four times smaller (20% mzML size) instead of two times smaller (50%).

@Bouyssie2015 also found that mz5 was about 80% smaller than the mzML and on par with their mzDB format. They reported query times of around 30 seconds for a wide (5 Da) ion chromatogram extraction and found that mz5 was about 40 times slower and that mzML was 200 times slower, in contrast to our parser which was 10 times slower for mz5 - possibly due to the large EIC width. They also were able to report mzDB scan queries on par with mz5 which we were able to replicate if the pyopenms or pyteomics libraries were used for full scan queries and pymzml for MS/MS.

MzTree [@Handy2017] compared their SQLite-based system to mzML, mz5, and mzDB and reported high numbers for both EIC and RT range queries from mzML (4-1000 seconds) that were on par with the values we observed here. Their mzDB and mz5 results were unexpectedly comparable to each other at about 0.5 seconds per random EIC query, with the mzDB values equal to ours but the mz5 values much lower than our parser was able to obtain. Our disk size measurements also corresponded well with their size estimate of MzTree at approximately twice the size of the mzML, a surprising result given that our mzML file contained MS/MS information and theirs appearing to be full-scan only. Their mz5 files were much larger (80% mzML size instead of our 20%) and their mzDB much smaller (20% mzML size instead of our 110%).

The mzMD format [@Yang2022] appears to be a thin wrapper around the MzTree format that applies a different philosophy for data subsetting and summarization. They report EIC queries in the 50 millisecond range, very similar to the values we obtained for the mzMD file type. They also estimate file size to be approximately 28 bytes per *m/z*/intensity tuple for a total size of 100MB in our test file which agrees reasonably well with our 72MB actual measurement. Their comparison to MzTree also agrees with our results as they report slightly larger disk usage and slightly better performance.

The mzMLb group [@Bhamber2021] reports only info for spectrum access at approximately 15ms/scan which agrees with our 5-100ms/scan estimates only if the data is loaded ahead of time. They perform extensive comparisons to mzML at varying compression methods and levels but our default options of 1024 KB chunk sizes for the mzMLb file and zlib but not numpress compression for the mzML resulting in timing values very similar between mzML and mzMLb using the pyteomics library for access. They also compared to the mz5 file type and we are able to validate their results of the mzML+zlib occupying significantly more space, though our mz5 parser outperformed theirs for full scan and MS/MS data by two orders of magnitude in time.

We were also unable to test several other recent and promising formats. Aird does not report full query times for any of the metrics reviewed here, though they claim their StackZDPD algorithm [@Wang2022] can improve decompression speed by three times and that the file size is 54% of the vendor file.[@Lu2022] Similarly, the toffee format for time-of-flight DIA data reports sizes about equal to vendor or 60% of centroid mzml + numpress with query speeds 4 times faster for scans (spectrum-centric, 2 seconds for mzML, 0.5 for toffee) and 100 times faster for chroms (peptide-centric, 168 seconds for mzML, 1.8 for toffee).[@Tully2020] The Unified Ion Mobility Format (UIMF, @Beagley2009) from Pacific Northwest National Laboratory format did not report direct comparisons to any of the available formats and thus we must remain unclear on its performance capabilities.

### Fundamental inefficiencies in existing mass-spectrometry formats

We identified several fundamental inefficiencies when writing the parsers. First, scan metadata that was encoded within a scan instead of in a separate unit required looping over each scan to see whether it contained the information requested. Scan number, MS level, and retention time were all useful bits of information that could be included in a file header or footer to relate the three to the data location within the file and allow index use instead of looping over every scan. Second, needing to decode or parse a compressed *m/z*/intensity array in each scan introduced an additional overhead that was especially punishing during ion chromatogram extraction and MS/MS search. While the *m/z* and intensity tuples are an obvious candidate for data compression, this penalty should be of significant concern to engineers. Third, looping over files is inherently slow and introduces additional complexity relative to a single unified database that encodes filename or sample ID as an additional column. A particular strength of the database system we propose is its inherent support for multi-file systems, while all other methods require looping over files.

The problems above highlight an important distinction between data *access* and data *search* that has been largely overlooked in our opinion. While HDF5 files or scan indexes excel at improving data access, they assume that the location of the data is known in advance and can be skipped to via bitwise offsets. If a search is required, however, this advantage is fully negated because each bit of information must be queried anyway. Finally, we must note that scan number is not inherently a useful bit of information. While we included it in our extraction metrics, it is entirely unclear when the scan number itself would be known in isolation. Additionally, this method often confuses the scan number with the scan's indexes in the data structure. Scan number is not always consecutive (i.e. during polarity switching, multi-experiment samples, or if any filtering is performed during processing), so even if the first or second item in the structure can be queried speedily this is no guarantee that the item will contain the information of interest.

### Leveraging robust, future-oriented software development with SQL

The proliferation of MS data storage formats and access algorithms illustrates the general dissatisfaction with existing alternatives to the vendor file or mzML. Formats that are faster to query or smaller on disk tend to be significantly more opaque, and those optimized for a particular method often fail to perform well on other metrics. This complexity is generally expected as optimization tends to require more complex data structures and assumptions about its use but it is not required if the complexity is outsourced to a robust and growing framework such as structured query language (SQL).

SQL is widely used for data processing outside of mass-spectrometry, though its adoption is increasing in recent years. Efforts like mzDB, the Pacific Marine Environment Laboratory's UIMF format [@Beagley2009], and the internals of MzTree hint at SQL's suitability for MS data storage. SQL backends for the next-generation R processing package Spectra now exist [@Rainer2022] and the development of MassQL [@Jarmusch2022] indicates a growing comfort with SQL syntax for downstream processing, though the language itself strives for human readability in simple queries. The searching and subsetting inherent to MS data exploration represent very simple queries in database space, agnostic to high-level programming language and rarely requiring more than a single line of code. Additionally, the extensive documentation that exists across the internet means that large language models such as ChatGPT are easily able to translate queries for those unfamiliar with SQL's syntax.

Just as the original database paper from @Codd1970 argued that the same problems were being solved over and over again, mass spectrometry data scientists are re-solving problems that have been more elegantly ironed out by dedicated teams in computer science and industry with much more extensive support. By leveraging existing optimizations in SQLite and DuckDB, we were able to create a highly performant system for storage of MS data that does not come with significant trade-offs between data extraction methods.

While SQLite is broadly used and its long history testifies to its continued utility, we can use even more modern database methods to improve further upon its analytical processing capacity. We tested both DuckDB the Apache Parquet data formats [@Raasveldt2019; @Vohra2016] and found that they both performed better than SQLite in disk usage and query speed. DuckDB in particular is nearly a drop-in replacement for SQLite in many cases that's been extensively optimized for MS-related queries given its online analytical processing (OLAP) structure. DuckDB provides automatic compression algorithms and uses zonemaps to create bounding boxes for each subset of data, bringing together existing optimizations from mz5 (delta encoding), mzDB (bounding boxes), and MzTree (axis-agnostic queries) at zero additional cost. Importantly, as with all databases, only the subset of interest needs to be written into memory, making the hardware requirements relatively lightweight.

Of course, to claim that existing frameworks should be discarded in favor of a novel method is to ignore decades of discussion and conversation. We acknowledge that our use case, that of largely exploratory and quality-control steps, is not a universal need and our lack of perfect metadata preservation in particular indicates that databases should become an auxiliary data structure alongside the vendor files or mzMLs, not substitute for them directly. Ultimately, the design decision for mass spectrometry data format will likely continue to be a point of contention and will result from a variety of factors, most crucially 1) initial vendor type, 2) programming language of the developer, 3) types of MS data included (e.g. full scan only versus MS/MS or metadata requirements), 4) whether the entire file will be processed or only a subset, and 5) how well a file type interfaces with downstream software. We intend to show with this manuscript that there is significant overlap between the goals of organizations much larger than any individual lab and that MS as a whole can benefit significantly from co-opting their development.

@@

We propose that a simple relational database is an intuitive and performant mass spectrometry (MS) data storage format. Tables containing fields that map directly to known MS concepts means that adoption is straightforward and facilitated by the widely-understood structured query language (SQL), reducing the code required to extract subsets of interest to a single line. We show that this structure can also take advantage of regular advancements in computer science by leveraging modern data formats such as DuckDB and Parquet to reduce the disk space required while improving access times by 1-2 orders of magnitude. We hope that widespread adoption of this format alongside the metadata-heavy vendor and mzML files will reduce the barriers to data access for mass spectrometrists and provide a consistent framework that covers a majority of the exploratory use cases.

[add R/Julia comparison?]
